 

### 2021-05-13 vertical-thermal-review.Rmd
total so far, 7962 words

**The biophysical environment across the vertical gradient from the forest floor to the the top of the canopy is in large part determined by the structure of the forest itself.** 
Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].

### Foliage density
**Canopy foliage, which is vertically structured in correspondence with the size, structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig. `r fig_NEON_vertical`).**
Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

@@

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017]. Light reaching forest floors that is incident to canopy tops ranges from ~1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997]. Sunflecks, caused by small canopy gaps and wind-induced canopy movements also contribute to the shaded forest regions that receive 50-80% of their light for photosynthesis during these short exposures [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. Consistent with these results, higher wind speeds at greater heights have been observed in both closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry season [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; @mcgregor_tree_nodate]. However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al. 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.

@@

**The combination of biophysical variables reviewed above drive leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn has a strong influence on leaf metabolism (see following section).**
Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum $T_{leaf}$ in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. As a result, leaves can be cooler than air at high temperature and warmer than air at low temperature ('limited homeothermy') or can be drastically warmer than $T_{air}$ ('megathermy'). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ ('poikilothermy') [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

$T_{leaf}$ itself is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`). The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. 2; @michaletzPlantThermoregulationEnergetics2015].

@@

Leaf thermoregulation is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. Leaves can regulate their temperature by dissapating heat through stomatal conductance ($g_s$), even in drastic biophysical conditions [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

Incoming solar radiation is a strong driver of $T_{leaf}$, relative to $T_{air}$. With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

Greater wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, canopy-leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al. 2018]. At low wind conditions, small $g_a$ to heat transfer can increase $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017] where boundary layer resistance increases with thickness of air layer at the leaf surface. This tends to be greater in buffered leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
While stomatal conductance is inherently greater in exposed leaves than buffered shaded leaves, factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--are also greater with exposure. As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under increased $T_{air}$ or irradiance.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.**  Smaller leaves remain within a few degrees of $T_{air}$,  *with coolest leaves at intermediate sizes (~10mm)* due to higher $g_{a}$ to heat transfer [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017]. Whereas, large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed  [@leighInfluenceLeafSize2017; @fauset_differences_2018].

### In-situ observations
**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Canopy buffering capacity largely influences vertical $T_{leaf}$ dynamics. Particularly, in open-forests with lower buffering capacity, $T_{leaf}$ is greater in lower-canopy than upper canopy, reflecting microclimate conditions of vertical air mixing, light permeating through canopy layers, still air in the lower canopy and heating from the ground, [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. Greater $RH$ in inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies due to greater heat and water stress from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

@@

**Leaf thermoregulation largely depends leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016]. Therefore, are critical in maintaining desirable $T_{leaf}$ and in shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004]. Degrees of direct to diffused light and resultant water demands shape sun vs. shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. Therefore, taller tree canopy-leaves dramatically vary from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). 
While leaf traits vary primarily on the axis of light, at greater tree-heights, both theory and empirical evidence indicate that other factors-for example, tree-height or exposure to evaporative demand (Fig. `r fig_NEON_vertical`)- are also important in shaping at least a subset of traits.

@@

Disentangling the influences of hydraulic constraints, light and other biophysical drivers that vary with height (Fig. `r fig_NEON_vertical`) on leaf traits remains an important area for research, but is beyond the scope of the current review. Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests. For the purposes of interpreting the results presented here, we note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.

@@

Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases. Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, shaded leaves invest more in $SLA$ for light capture than in heat dissapation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Additionally, under homogeneous light conditions, sun-leaves in different canopy orientations are shaped on the basis of $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

Nitrogen along the vertical gradient, is constantly re-mobilized from senescensed to growing leaves [@niinemetsPhotosynthesisResourceDistribution2007a] where $N_{area}$, with LMA, leaf soluble sugars, starch and non-structural carbons are greater [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014]. Additionally, $N_{area}$ optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].

@@

Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--plays a role in converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], increases proportional light gradients.

@@

Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and stomatal conductance with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies largely with inherent specie-specific trait differences such as leaf shape, size, $gs$ levels and therefore influences individual $T_{leafs}$. As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

$T_{leaf}$ beyond photosynthetic $T_{opt}$ declines photosynthesis due it's peaked response to temperature. Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020]. Leaves in tropical upper canopies have shown to experience in-situ $T_{air}$ well above $40-60{^\circ}C$ on a regular basis. If $T_{leaf}$ is not regulated or adapated, heat stress can loss of canopy leaves [**REF**]

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].

>>>>

**The biophysical environment across the vertical gradient from the forest floor to the top of the canopy is in large part determined by the structure of the forest itself.** 
Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].

### Foliage density
**Canopy foliage, acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions and is critical in influencing-and is largely influenced by-the vertical biophysical gradient.(Fig. `r fig_NEON_vertical`).**The distribution of this foliage is structured in correspondence with the size, and varies across tree structures and forests.  
Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

@@

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density. This gradient is most pronounced under mixed and broad-leaf forests where light reaching forest floors, incident to the canopy-top range from ~1-2% to 4-5%, respectively [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997]. Sunflecks, caused by small canopy gaps and wind-induced canopy movements are an important light-source for shaded forest regions that rely on sunflecks for photosynthesis [@waySunflecksTreesForests2012]. In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height. With daily maximum wind speeds averaging up to **#** $m s^-1^$ higher at the top of the NEON vertical profile than at the bottom

@@

The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, more than open-forests, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry season [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; @mcgregor_tree_nodate]. However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss in results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al. 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate].

@@

**Vertical biophysical drivers reviewed above influence heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn shapes leaf metabolism (see following section).**
Along the complex biophysical dynamics, leaves adjust to their environment by maintaining optimum $T_{leaf}$ for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. Therefore, leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles  [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`) where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. 2; @michaletzPlantThermoregulationEnergetics2015].

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because stomatal conductance ($g_s$) facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling [Drake et al. 2018] but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves than buffered shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$ [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].

### In-situ observations
**The principles outlined above, vertical microclimate profile (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and evapotranspiration (where is this discussed? and do we say anything about evaporation?) together shape in-situ $T_{leaf}$ observations (Fig. `r fig_leaf_T`).**

Particularly, in open-forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016]. Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. 
Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

@@

**Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016]. Therefore, are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs. shade leaves, and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). 
But at greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig. `r fig_NEON_vertical`)--also play a role in shaping leaf traits

@@

However, disentangling tree-height related biological and biophysical drivers (Fig. `r fig_NEON_vertical`) on leaf traits remains an important area for research, which is sparse, but is beyond the scope of the current review. Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests. We note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.

@@

Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where  sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with adaptive strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance.On the other hand, shaded leaves invest more in light capture than in heat dissipation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous light conditions, sun-leaves in different canopy orientations are shaped more based on $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].

@@

The distribution of biochemical traits enable photosynthesis and metabolic processes. Particularly $N_{area}$, is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a] and is optmized to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].

@@

Xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--are greater in sun-leaves where they enable heat and light tolerance by converting excess light to heat dissapation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. $VAZ$, and in general biochemical, constraints can result in lower temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]

@@

Across tree-size ontogeny of the same species, while fundamental trait plasticity to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and $gs$ with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies greatly with inherent specie-specific trait differences. As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

$T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature. Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020]. Leaves in tropical upper canopies have shown to experience in-situ $T_{air}$ well above $40-60{^\circ}C$ on a regular basis. If $T_{leaf}$ is not regulated or adapated, heat stress can loss of canopy leaves [**REF**]

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
